---
title: "ConfoundR Vignette"
author: "John W. Jackson, Erin Schnellinger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ConfoundR Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(confoundr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

##Example code

Below, we provide an example in which we apply Diagnostic 3 to an artificial data set to 
illustrate covariate balance for a time-varying exposure without censoring. 

```{r}
##########################
##LOAD DATA AND PACKAGES##
##########################

#NOTE THAT THIS CODE REQUIRES R VERSION 3.1.2 OR HIGHER 
library(magrittr) #last tested on magrittr v1.5
library(tidyr)    #last tested on tidyr v0.6.0
library(dplyr)    #last tested on dplyr v0.5.0
library(ggplot2)  #last tested on ggplot2 v2.1.0
library(grid)     #this comes with base R
library(gridExtra)#last tested on gridExtra v2.0.1
library(scales)   #last tested on ggplot2 v0.4.0
library(Rmpfr)    #last tested on Rmpfr v0.6-0 
library(broom)    #last tested on broom v0.4.1

# Load the example data
data("example_sml")


#####################################################################
##Example: Diagnostic 3 for a time-varying exposure without censoring
#####################################################################

#PRELIMINARY STEP: MAKE EXPOSURE HISTORY

# Remove history for illustration. Note that the input data is already in wide format.
drops <- c("h_0", "h_1", "h_2")
mydata <- example_sml[ , !(names(example_sml) %in% drops)]

mydata.history <- makehistory.one(input=mydata,
                                  id="id",
                                  exposure="a",
                                  name.history="h",
                                  times=c(0,1,2))

#STEP 1: RESTRUCTURE THE DATA
mydata.long <- lengthen(
  input=example_sml, #mydata.history,
  id="id",
  diagnostic=3,
  censoring="no",
  times.exposure=c(0,1,2),
  times.covariate=c(0,1,2),
  exposure="a",
  temporal.covariate=c("l","m","o"),
  static.covariate=c("n","p"),
  history="h",
  weight.exposure="wax"
)

#example of how to remove relative covariate history
mydata.long.omit <- omit.history(input=mydata.long,
  omission="relative",
  covariate.name=c("l","m","o"),
  distance=1
  )


#STEP 2: CREATE BALANCE TABLE 
mytable <- balance (
input=mydata.long.omit,
diagnostic=3,
approach="weight", 
censoring="no",
scope="all",
times.exposure=c(0,1,2),
times.covariate=c(0,1,2),
exposure="a",
history="h",
weight.exposure="wax",
ignore.missing.metric="no",
sort.order= c("l","m","o","n","p")
)

#STEP 3: PLOT BALANCE METRIC
myplot <- makeplot (
input=mytable,	
diagnostic	=3,
approach="weight",
#censoring="no",
scope="all",
metric="SMD"
)
#The following formatting arguments for makeplot() are optional (defaults shown).

#label.exposure="A",							#exposure label
#label.covariate="C",							#covariate label
#lbound=-1,									    	#lower bound for x-axis
#ubound=1,										  	#upper bound for x-axis
#ratio=2,										    	#plot aspect ratio
#text.axis.title=8,								#title font size
#text.axis.y=6.5,									#y-axis (covariate names) font size
#text.axis.x=6.5,									#x-axis font size	
#text.strip.y=10,									#row panel label font size
#text.strip.x=10,									#column panel label font size
#point.size=.75,									#dot size
#zeroline.size=.1,								#thickness of zero line on x-axis
#refline.size=.1,									#thickness of reference line on x-axis
#refline.limit.a=-.25,						#location for reference line 1 on x-axis
#refline.limit.b=0.25,						#location for reference line 2 on x-axis
#panel.margin.size=.75,						#space between panels
#axis.title="Mean Difference",    #or "Standardized Mean Difference" (x-axis title)
#label.width=15									  #width of panel label text (before wrapping text)

#Display the plot
myplot

#STEP 4: SAVE BALANCE TABLE AND PLOT
#write.csv(mytable,paste(path,"mytable.csv",sep=""))
#ggsave(filename=paste(path,"myplot.pdf",sep=""))

```


Now, we illustrate how Diagnostic 1 can be implemented using a regression approach. 

```{r}
#################################################
##Example of Regression Approach for Diagnostic 1
#################################################

library(broom) #need for tidy()

#create long dataset
mydata.long <- lengthen(input=mydata.history,
  diagnostic=1,
  censoring="no",
  id="id",
  times.exposure=c(0,1,2),
  times.covariate=c(0,1,2),
  exposure="a",
  temporal.covariate=c("l","m","n","o","p"),
  history="h"
)         

##MAKE BALANCE TABLE USING REGRESSION##

#create balance table
mydata.long.reg <- mutate(mydata.long,time=time.exposure,distance=time.exposure-time.covariate,history=h)
output <- mydata.long.reg %>% 
  group_by(name.cov) %>% #note, you can include other stratifying variables here or in the model
    filter(time.exposure>=time.covariate) %>%
      do(tidy(lm(formula=value.cov~a+time+distance+history,.))) %>% #same model form used for every covariate
        filter(term=="a1") %>% ungroup()

table.reg <- output %>% 
               select(name.cov,estimate) %>% 
                 rename_("D"="estimate")

print(table.reg)
#write.csv(table.reg,paste(path,"table_regression.csv"))
#NOTE: This code applies the same model parameterization for each covariate (relying on a strong assumption).
```

Finally, we compare the regression approach to the direct calculation and standardization methods implemented by confoundR.

```{r}
### COMPARE THAT TO A DIRECT CALCULATION & STANDARDIZATION ###

table.std <- balance(input=mydata.long,
  diagnostic=1,
  approach="none",
  censoring="no",
  scope="average",
  average.over="distance",
  ignore.missing.metric="no",
  times.exposure=c(0,1,2),
  times.covariate=c(0,1,2),
  exposure="a",
  history="h"
) 

print(table.std)
#write.csv(table.std,paste(path,"table_standardization.csv"))

```

